---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
## `glmmsr`: fit GLMMs with sequential reduction

Generalized linear mixed models (GLMMs) are an important and widely-used model 
class. In R, we can fit these models with the 
[`lme4`](https://github.com/lme4/lme4) package, but there are some
limitations. First, except in very simple cases, `lme4` uses a Laplace 
approximation to the likelihood for inference, which may be of poor quality in 
some cases. Second, it is difficult to fit some GLMMs, such as pairwise 
comparison models, with `lme4`. 

The `glmmsr` package aims to offer progress on both of these problems. 

The sequential reduction approximation to the likelihood 
is controlled by a parameter (`k`) 
which allows the user to trade-off accuracy of the approximation against 
computation time. 

`glmmsr` also provides an extended interface to `lme4`. This
interface allows easy fitting of pairwise comparison and many other interesting 
models, which are difficult to fit directly with `lme4`.

The vignette gives more details and examples of how to use 
the sequential reduction approximation and the subformula 
interface.

## Installing `glmmsr`

You will need the [devtools](https://github.com/hadley/devtools) package.

Then run
```{r, eval = FALSE}
devtools::install_github("heogden/glmmsr", build_vignettes = TRUE)
```

```
## Documentation

To view the vignette for `glmmsr`, use
```{r, eval = FALSE}
browseVignettes("glmmsr")
```
or see [here](http://htmlpreview.github.io/?https://github.com/heogden/glmmsr/blob/master/inst/doc/glmmsr-vignette.html).

